Forecasting the 2024 US Presidential Election

### 2024-10-27 paper/paper.qmd
added timeline data

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(dplyr)
library(knitr)
library(lubridate)
library(ggplot2)
library(ggcorrplot)
library(kableExtra)

@@

set.seed(754)
poll_data <- read_csv(here::here(
  "data/02-analysis_data/organized_state_poll_data.csv"))

@@

# Function to create the plot with y-axis from 0 to 100 and data labels
create_comparison_plot <- function(plot_data) {
  ggplot(plot_data, aes(
    x = reorder(Variable, -Mean), 
    y = Mean, 
    fill = Group, 
    label = round(Mean, 1)
  )) +
    geom_col(position = position_dodge(width = 0.9)) +
    geom_text(
      position = position_dodge(width = 0.9), 
      vjust = -0.3, 
      size = 3
    ) +
    scale_fill_manual(
      values = c("Blue States" = "red", "Red States" = "blue")  # Reversed colors
    ) +
    labs(
      title = "Demographic Differences Between Top Red and Blue States",
      x = "Demographic Variable",
      y = "Mean Percentage",
      fill = "State Group"
    ) +
    scale_y_continuous(limits = c(0, 100)) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 4)
    )
}

@@

# Run the analysis
analysis_results <- analyze_state_differences(poll_summary, top_n_states = 1, top_n_vars = 6)

@@

column_refs <- list(
  democrat = "democrat_pct",
  republican = "republican_pct",
  highly_educated = "What is the highest level of education you have attained? College graduate",
  lower_educated = "What is the highest level of education you have attained? High school or less",
  female = "Can you please tell me your gender? Female",
  male = "Can you please tell me your gender? Male",
  non_binary_other = "Can you please tell me your gender? Nonbinary or other",
  caucasian = "For statistical purposes only, can you please tell me your ethnicity? White or Caucasian",
  hispanic_latino = "For statistical purposes only, can you please tell me your ethnicity? Hispanic or Latino of any race",
  vote_trump = "Who did you vote for in the 2020 election? Donald Trump",
  vote_biden = "Who did you vote for in the 2020 election? Joe Biden",
  biden_approve = "Do you approve or disapprove of the job Joe Biden is doing as President? Approve",
  biden_disapprove = "Do you approve or disapprove of the job Joe Biden is doing as President? Disapprove",
  young_age_18_29 = "What is your age range? 18-29 years",
  young_age_30_39 = "What is your age range? 30-39 years",
  young_age_40_49 = "What is your age range? 40-49 years",
  older_age_50_59 = "What is your age range? 50-59 years",
  older_age_60_69 = "What is your age range? 60-69 years",
  older_age_70_more = "What is your age range? 70 or more years"
)
```

@@

This analysis offers valuable insights into how specific demographic factors shape the political landscape in different states, allowing us to better predict electoral outcomes based on shifting demographic trends.

>>>>

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(dplyr)
library(knitr)
library(lubridate)
library(ggplot2)
library(ggcorrplot)
library(kableExtra)
library(arrow)
library(zoo)
```

@@

set.seed(754)
poll_data <- read_parquet(here::here(
  "data/02-analysis_data/summarized_state_poll_data.parquet"))
time_data <- read_parquet(here::here(
  "data/02-analysis_data/timeline_data.parquet"))

@@

# Function to create the plot with y-axis from 0 to 100 and data labels
create_comparison_plot <- function(plot_data) {
  ggplot(plot_data, aes(
    x = reorder(Variable, -Mean), 
    y = Mean, 
    fill = Group, 
    label = round(Mean, 1)
  )) +
    geom_col(position = position_dodge(width = 0.9)) +
    geom_text(
      position = position_dodge(width = 0.9), 
      vjust = -0.3, 
      size = 3
    ) +
    scale_fill_manual(
      values = c("Blue States" = "blue", "Red States" = "red")
    ) +
    labs(
      title = "Demographic Differences Between Top Red and Blue States",
      x = "Demographic Variable",
      y = "Mean Percentage",
      fill = "State Group"
    ) +
    scale_y_continuous(limits = c(0, 100)) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 7)
    )
}

@@

# Run the analysis
analysis_results <- analyze_state_differences(poll_summary, top_n_states = 1, top_n_vars = 5)

@@

column_refs <- list(
  democrat = "democrat_pct",
  republican = "republican_pct",
  highly_educated = "What is the highest level of education you have attained? College graduate",
  lower_educated = "What is the highest level of education you have attained? High school or less",
  female = "Can you please tell me your gender? Female",
  male = "Can you please tell me your gender? Male",
  non_binary_other = "Can you please tell me your gender? Nonbinary or other",
  caucasian = "For statistical purposes only, can you please tell me your ethnicity? White or Caucasian",
  hispanic_latino = "For statistical purposes only, can you please tell me your ethnicity? Hispanic or Latino of any race",
  vote_trump = "Who did you vote for in the 2020 election? Donald Trump",
  vote_biden = "Who did you vote for in the 2020 election? Joe Biden",
  biden_approve = "Do you approve or disapprove of the job Joe Biden is doing as President? Approve",
  biden_disapprove = "Do you approve or disapprove of the job Joe Biden is doing as President? Disapprove",
)
```

@@

```{r}
# Convert `end_date` to Date format and ensure numeric columns
time_data <- time_data %>%
  mutate(end_date = as.Date(end_date, format = "%m/%d/%y"),
         pct = as.numeric(pct), 
         `Do you approve or disapprove of the job Joe Biden is doing as President? Approve` = 
           as.numeric(`Do you approve or disapprove of the job Joe Biden is doing as President? Approve`))

# Apply a 14-day rolling average (increase smoothing) to each group
time_data <- time_data %>%
  group_by(party) %>%
  mutate(pct_smooth = rollapply(pct, width = 7, FUN = mean, align = "right", fill = NA)) %>%
  ungroup() %>%
  mutate(biden_approval_smooth = rollapply(`Do you approve or disapprove of the job Joe Biden is doing as President? Approve`, 
                                           width = 7, FUN = mean, align = "right", fill = NA))

# Separate data for Democrats, Republicans, and Biden Approval
# Filter data for Democrats where state is NA
democrat_data <- time_data %>%
  filter(party == "DEM" & is.na(state))

# Filter data for Republicans where state is NA
republican_data <- time_data %>%
  filter(party == "REP" & is.na(state))

biden_approval <- time_data %>%
  filter(is.na(state)) %>%
  select(end_date, biden_approval_smooth)

# Create the time plot with ggplot2
ggplot() +
  # Plot Democrats' smoothed pct over time (blue)
  geom_line(data = democrat_data, aes(x = end_date, y = pct_smooth, color = "Democrats"), size = 1) +

  # Plot Republicans' smoothed pct over time (red)
  geom_line(data = republican_data, aes(x = end_date, y = pct_smooth, color = "Republicans"), size = 1) +

  # Plot Biden's smoothed Approval over time (green)
  geom_line(data = biden_approval, aes(x = end_date, y = biden_approval_smooth, color = "Biden Approval"), size = 1) +

  # Customize labels, theme, colors, and y-axis limits
  labs(title = "Democrat vs. Republican and Biden Approval Over Time (Smoothed)",
       x = "Date",
       y = "Percentage (%)",
       color = "Legend") +
  scale_color_manual(values = c("Democrats" = "blue", "Republicans" = "red", "Biden Approval" = "lightblue")) +
  ylim(25, 55) + # Set y-axis range from 25 to 55
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
This analysis offers valuable insights into how specific demographic factors shape the political landscape in different states, allowing us to better predict electoral outcomes based on shifting demographic trends.
